---
title: "PON Governance LLM Pilot"
subtitle: "Reconstructing the 2024 SBS protocol and comparing model generations"
editor: visual
---

```{r}
#| label: setup
#| include: false
library(dplyr)
library(knitr)
library(readr)
library(yaml)

cfg <- yaml::read_yaml("config/config.yml")
summary_path <- "derived/sbs_model_metrics_overall.csv"
summary_ready <- file.exists(summary_path)
if (summary_ready) {
  comparison_models <- purrr::map_dfr(
    cfg$comparison$comparison_models,
    ~ tibble::as_tibble(.x)
  ) |>
    dplyr::transmute(model_id = as.character(.data$model_id), model = as.character(.data$label))
  expected_model_ids <- c(cfg$legacy_protocol$model_alias, comparison_models$model_id)
  summary_metrics <- readr::read_csv(summary_path, show_col_types = FALSE)
  summary_ready <- setequal(summary_metrics$model_id, expected_model_ids)
}
```

## Repository purpose

This repository contains the computational materials associated with the manuscript *Institutional Rules, Boundary Objects, and LLMs — A proof-of-concept for large-N research on the governance of purpose-oriented networks*.

The revised principal analysis asks:

> **How do contemporary OpenAI models compare with the frozen 2024 GPT-4 Turbo responses when all are evaluated under the same reconstructed step-by-step (SBS) protocol?**

The historical benchmark is the preserved Dec. 24, 2024 `*_sbs.txt` output. The contemporary arms are the pinned GPT-5.4 nano snapshot (`gpt-5.4-nano-2026-03-17`) and GPT-5.6 Luna (`gpt-5.6-luna`). OpenAI's Luna documentation was checked on 2026-08-31 and did not expose a dated Luna snapshot, so `gpt-5.6-luna` is recorded as an undated model ID together with execution-time provenance. The original 2024 outputs are never regenerated or overwritten.

## Study cases

The strict comparison uses **IRG, EUHC, cOAlition S (COAS), CAN, and CEER**. BEREC remains part of the archival corpus but is handled separately because its preserved Dec. 23 output predates the final Dec. 24 request implementation.

Dedicated BEREC replication uses `BO1B.txt`, not `BO1A.txt`, and is excluded from the five-PON aggregate analysis.

## Two linked reproductions

The repository separates two tasks:

1. **Manuscript Table A5 replication.** HBM, ASC, JNG, and the frozen legacy SBS LLM reconstruct the historical four-coder agreement table over the manuscript's 14 prompts.
2. **Extended SBS model comparison.** The same human data and frozen legacy SBS outputs are compared with GPT-5.4 nano and GPT-5.6 Luna SBS runs over the preserved 15-question computational instrument.

Project-facing outputs use explicit agreement labels such as `3 TRUE / 1 FALSE`, and report directly whether each LLM agrees with the three-human majority vote.

## Principal comparison result

The table below is generated directly from the stored three-model evaluation outputs. It reports agreement with the confirmed three-human majority reference over the same 75 PON-question classifications for each model arm; no result is manually transcribed into this page.

```{r}
#| label: tbl-principal-summary
#| tbl-cap: "Principal SBS agreement with the confirmed human reference"
if (summary_ready) {
  summary_metrics |>
    dplyr::transmute(
      Model = .data$model,
      Decisions = .data$n,
      Agreement = scales::percent(.data$accuracy, accuracy = 0.1)
    ) |>
    knitr::kable()
} else {
  tibble::tribble(
    ~Model, ~Status,
    "GPT-4 Turbo", "Frozen benchmark available",
    "GPT-5.4 nano", "Completed run available",
    "GPT-5.6 Luna", "Completed run available; rebuild derived metrics"
  ) |>
    knitr::kable()
}
```

These figures describe this fixed five-PON, 15-question comparison and should not be interpreted as a general benchmark of any model family. See [SBS Model Comparison](reports/model-comparison.qmd) for the full breakdown.

## Table A5 provenance

The surviving data reproduce 67 of 70 Table A5 cells. Three manuscript cells report `2-2` where the raw human + frozen LLM responses produce a 3-to-1 split: **IRG prompt 3, cOAlition S prompt 8, and CEER prompt 6**. These are retained as documented Table A5 discrepancies relative to the surviving coder-level data; no underlying response is altered.

See [2024 Replication](reports/results.qmd) and [Table A5 replication provenance](provenance/TABLE_A5_REPLICATION.md).

## Reproducibility architecture

The repository separates:

- frozen 2024 evidence in root-level `output/` files;
- new SBS model runs in timestamped directories under `output/current/`, with completed runs explicitly marked by `run_manifest.json` and partial/failed runs retained as `run_failed.json`;
- dedicated BEREC runs under `output/berec_replication/`;
- archival improved-protocol materials under `output/improved/`; and
- derived tables and figures under `derived/`.

For execution instructions and API safeguards, see `README.md`.
