---
title: "BEREC Reconstruction"
---

## Scope

BEREC is analyzed separately from the principal five-PON model comparison.

The main comparison uses IRG, EUHC, cOAlition S, CAN, and CEER under a common reconstructed step-by-step (SBS) protocol. BEREC presents a different provenance problem: a preserved historical output exists, but the exact historical interaction cannot be reconstructed with sufficient confidence to treat a new run as an exact replication.

The BEREC analysis therefore compares three distinct conditions without treating any of them as ground truth.

## Three BEREC conditions

1. **Preserved historical output.** The surviving December 23, 2024 BEREC classification artifact is preserved unchanged in `output/BO1A-2024-12-23-55_cat.txt`.

2. **Best-evidenced December 23 reconstruction.** The reconstructed historical-family condition uses `BO1B.txt`, the best-evidenced reconstructed December 23 prompt configuration, and the pinned `gpt-4-turbo-2024-04-09` model. This is explicitly a reconstruction, **not an exact-replication claim**.

3. **Retrospective December 24 SBS application with GPT-5.4 nano.** The recovered December 24 SBS protocol is applied to `BO1B.txt` using the pinned `gpt-5.4-nano-2026-03-17` snapshot.

The derived comparison tables below are generated locally from those artifacts. No API calls are made during this analysis.

```{r}
#| echo: false
#| warning: false
#| message: false

library(yaml)
cfg <- yaml::read_yaml("config/config.yml")
publication_excluded_question_ids <- as.integer(unlist(cfg$publication$excluded_question_ids))

berec_summary <- utils::read.csv(
  "derived/berec_three_way_summary.csv",
  stringsAsFactors = FALSE,
  check.names = FALSE
)

berec_comparison <- utils::read.csv(
  "derived/berec_three_way_comparison.csv",
  stringsAsFactors = FALSE,
  check.names = FALSE
)

berec_sources <- utils::read.csv(
  "derived/berec_sources.csv",
  stringsAsFactors = FALSE,
  check.names = FALSE
)

# Validate the complete artifact first, then apply the publication policy.
if (nrow(berec_comparison) != 15L || anyDuplicated(berec_comparison$question_id)) {
  stop("Unexpected BEREC computational comparison cardinality")
}
berec_comparison <- berec_comparison[
  !berec_comparison$question_id %in% publication_excluded_question_ids,
  ,
  drop = FALSE
]
berec_comparison$publication_question_id <- berec_comparison$question_id - vapply(
  berec_comparison$question_id,
  function(id) sum(publication_excluded_question_ids < id),
  integer(1)
)

pairwise_summary <- function(a, b, label) {
  same <- sum(a == b)
  different <- sum(a != b)
  data.frame(
    comparison = label, questions = length(a), same = same, different = different,
    agreement_rate = same / length(a),
    true_to_false = sum(a == "TRUE" & b == "FALSE"),
    false_to_true = sum(a == "FALSE" & b == "TRUE"),
    check.names = FALSE
  )
}

berec_summary <- rbind(
  pairwise_summary(berec_comparison$preserved_historical, berec_comparison$dec23_reconstructed_gpt4, "Preserved historical vs. reconstructed GPT-4 Turbo"),
  pairwise_summary(berec_comparison$preserved_historical, berec_comparison$dec24_sbs_gpt54_nano, "Preserved historical vs. GPT-5.4 nano SBS"),
  pairwise_summary(berec_comparison$dec23_reconstructed_gpt4, berec_comparison$dec24_sbs_gpt54_nano, "Reconstructed GPT-4 Turbo vs. GPT-5.4 nano SBS")
)

```

## Agreement across conditions

```{r}
#| label: tbl-berec-pairwise-agreement
#| echo: false
#| tbl-cap: "Pairwise agreement among the three BEREC conditions"

summary_display <- berec_summary

summary_display$agreement_rate <- sprintf(
  "%.1f%%",
  100 * summary_display$agreement_rate
)

names(summary_display) <- c(
  "Comparison",
  "Questions",
  "Same",
  "Different",
  "Agreement",
  "TRUE → FALSE",
  "FALSE → TRUE"
)

knitr::kable(
  summary_display,
  row.names = FALSE
)
```

The table reports classification agreement across the published question subset. These percentages describe **classification agreement**, not accuracy; the preserved historical artifact is an observed historical output, not a validated ground-truth coding of BEREC.

## Question-level comparison

```{r}
#| label: tbl-berec-question-comparison
#| echo: false
#| tbl-cap: "Question-level comparison of preserved and reconstructed BEREC classifications"

question_display <- berec_comparison[
  ,
  c(
    "publication_question_id",
    "question",
    "preserved_historical",
    "dec23_reconstructed_gpt4",
    "dec24_sbs_gpt54_nano"
  )
]

names(question_display) <- c(
  "Q",
  "Question",
  "Preserved historical",
  "Dec. 23 reconstructed GPT-4",
  "Dec. 24 SBS GPT-5.4 nano"
)

knitr::kable(
  question_display,
  row.names = FALSE
)
```

## Interpretation

The BEREC reconstruction provides evidence about the recoverability of the historical classification pattern across the published question subset. The comparison remains descriptive: the surviving evidence does not establish every element of the original BEREC request strongly enough to support an exact-replication claim, and differences between reconstructed conditions cannot be attributed exclusively to model change when protocol conditions also differ.

For this reason, BEREC remains an **auxiliary provenance and reconstruction analysis** and is not added as a sixth organization to the principal five-PON model comparison.


## Provenance

```{r}
#| label: tbl-berec-sources
#| echo: false
#| tbl-cap: "Sources used in the BEREC reconstruction analysis"

source_display <- berec_sources

names(source_display) <- c(
  "Condition",
  "Artifact",
  "Model",
  "Protocol",
  "Claim"
)

knitr::kable(
  source_display,
  row.names = FALSE
)
```

The preserved historical artifact remains read-only. Both reconstruction runs were written to new timestamped directories under `output/berec_replication/`.

The numerical results on this page are read from:

- `derived/berec_three_way_comparison.csv`
- `derived/berec_three_way_summary.csv`
- `derived/berec_sources.csv`

These files are generated by:

```text
python scripts/08_compare_berec.py
```

That script performs local comparison only and makes **no API calls**.
