---
title: "Methods"
subtitle: "A zero-shot 3 × 2 model-by-protocol comparison"
editor: visual
---

```{r}
#| label: setup
#| include: false
library(dplyr)
library(knitr)
library(readr)
library(yaml)

cfg <- yaml::read_yaml("config/config.yml")
questions <- readr::read_csv(
  cfg$project$questions_file,
  show_col_types = FALSE
)
```

## Study objective

The pilot asks whether LLMs can extract governance features from textual boundary objects associated with purpose-oriented networks (PONs). Each document is coded against 15 ordered yes/no governance questions.

The extension uses a **zero-shot 3 × 2 design** that separates two sources of variation:

1. **Model:** GPT-4 Turbo, GPT-5.4 nano, or GPT-5.4 mini.
2. **Protocol:** the recovered 2024 prompt or a concise improved prompt.

This yields six scenarios while keeping the five study documents, 15 questions, endpoint, message roles, and binary task fixed.

## Cases and source documents

```{r}
#| label: tbl-cases
#| tbl-cap: "PONs and boundary-object files retained in the reconstructed project"

tibble::tribble(
  ~PON, ~File,
  "IRG", "BO2A.txt",
  "EUHC", "BO3A.txt",
  "COAS", "BO4A.txt",
  "CAN", "BO5A.txt",
  "CEER", "BO6A.txt"
) |>
  knitr::kable()
```

The comparative experiment uses IRG, EUHC, COAS, CAN, and CEER. These five cases have recovered human-reference vectors tied to the final Dec. 24 computational protocol.

## Binary coding instrument

```{r}
#| label: tbl-questions
#| tbl-cap: "Binary governance questions in the preserved computational instrument"

questions |>
  knitr::kable(
    col.names = c("ID", "Question")
  )
```

::: {.callout-note}
The manuscript later reports 14 prompts. The preserved computational instrument contains one additional working-groups item. It remains in all six scenarios so the comparison does not change the outcome space retrospectively.
:::

## Zero-shot only

The analytical design no longer uses the historical `Let's think step by step` condition. Those files remain preserved as archival evidence, but they are excluded from evaluation, plots, tables, and new API runs. Removing that condition makes the comparison easier to interpret: each model/protocol cell differs on one intended dimension rather than mixing model generation with a weakly motivated prompting intervention.

## 2024 protocol

The final Dec. 24 implementation used Chat Completions with two messages:

1. a **system message** containing the policy-analyst instruction and ordered questions;
2. a **user message** containing the complete boundary-object text.

The request supplied only `model` and `messages` explicitly. No explicit temperature, top-p, token limit, seed, response format, or tools were supplied.

The preserved instruction is:

> You are a policy analyst focused on extracting organizational features from documents that regulate the organization and functioning of specific organizations. I will provide the documents and a list of yes-no questions. You will only return TRUE or FALSE responses according to the list of yes-no questions, including a row with the question itself. The output must be in CSV format.

The historical string concatenation, including its whitespace quirks, is retained for all three 2024-protocol cells.

## Improved protocol

The improved protocol keeps the same zero-shot task, documents, questions, Chat Completions endpoint, and message roles. The substantive treatment is the instruction text:

> Use only the supplied document. For each question, answer TRUE only if the text explicitly supports it; otherwise FALSE. Do not infer unstated rules or use outside knowledge. Return exactly 15 rows as question_id,response, preceded by that header. No other text.

The improvement is deliberately concise. It adds four things the old prompt left underspecified: a source boundary, an explicit TRUE/FALSE decision threshold, a prohibition on unstated inference, and an exact row-level output contract. These changes target both classification ambiguity and the serialization anomalies observed in the contemporary baseline run without introducing examples, chain-of-thought instructions, or a structured-output API parameter. The shorter wording also reduces prompt overhead while leaving the source document and questions unchanged.

All three improved-protocol model arms use the same 500-token maximum output allowance. The client expresses that ceiling with the model-appropriate Chat Completions parameter (`max_tokens` for GPT-4 Turbo and `max_completion_tokens` for GPT-5.4 nano and GPT-5.4 mini). The ceiling is an operational execution constraint rather than part of the classification prompt, is held constant across the three improved-protocol arms, and is recorded in the run metadata.

Before execution, the runner uses `tiktoken` to estimate the two-message request size for every PON. For the GPT-4 Turbo arm, estimated input plus the output cap must remain below 29,750 tokens: the observed 30,000-token account TPM ceiling less a 250-token safety margin. This local preflight was added after the API rejected the largest request at 30,012 requested tokens. Because Chat Completions framing can vary by serving implementation, the local count is treated as a conservative guardrail rather than an exact reconstruction of server-side accounting. A failed preflight creates no batch directory and makes no API call.

## Six scenarios

```{r}
#| label: tbl-design
#| tbl-cap: "Model-by-protocol scenarios"

tibble::tribble(
  ~Model, ~Protocol, ~Status,
  "GPT-4 Turbo", "2024 protocol", "Preserved 2024 zero-shot output",
  "GPT-5.4 nano", "2024 protocol", "Retained 2026 zero-shot run",
  "GPT-5.4 mini", "2024 protocol", "Retained 2026 zero-shot run",
  "GPT-4 Turbo", "Improved protocol", "2026 API run",
  "GPT-5.4 nano", "Improved protocol", "2026 API run",
  "GPT-5.4 mini", "Improved protocol", "2026 API run"
) |>
  knitr::kable()
```

The 2024 historical calls recorded the alias `gpt-4-turbo`, not a returned snapshot identifier. The improved-protocol arm therefore uses the pinned `gpt-4-turbo-2024-04-09` snapshot as the closest reproducible counterpart. The contemporary arms use pinned `gpt-5.4-nano-2026-03-17` and `gpt-5.4-mini-2026-03-17` snapshots under both protocols.

## BEREC exclusion

BEREC is retained only as archival provenance. Its preserved output predates the final Dec. 24 message-role implementation, and its boundary object is substantially larger than the other study documents. It is excluded from new default runs and from every comparative table, figure, aggregate statistic, and analytical denominator.

## Human-reference evaluation

The recovered computational workspace contains one human-reference vector per PON for IRG, EUHC, COAS, CAN, and CEER. All six scenarios are evaluated against these same 75 binary decisions. They should not be conflated with the later three-human-coder validation described in the manuscript.

## New-run provenance

Every new request saves the raw model output plus a JSON sidecar containing the requested and returned model IDs, completion ID, UTC request time, prompt and document hashes, usage data, system fingerprint when available, selected response headers, the explicit output-token limit when used, and the parameters deliberately left unspecified. Failed improved-protocol batches receive `batch_failed.json`; only fully completed batches receive `batch_manifest.json`. Single-model improved runs receive their own `run_manifest.json` only after completion.
