---
title: "Methods"
subtitle: "A fixed-protocol SBS model-generation comparison"
editor: visual
---

```{r}
#| label: setup
#| include: false
library(dplyr)
library(knitr)
library(readr)
library(yaml)

cfg <- yaml::read_yaml("config/config.yml")
questions <- readr::read_csv(cfg$project$questions_file, show_col_types = FALSE)
```

## Study objective

The pilot asks whether LLMs can extract governance features from textual boundary objects associated with purpose-oriented networks (PONs). The revised comparison holds the recovered manuscript-facing **step-by-step (SBS) protocol** fixed and varies model generation only.

The principal comparison is:

| Model | Protocol | Source |
|---|---|---|
| GPT-4 Turbo | Reconstructed Dec. 24, 2024 SBS | Frozen legacy output |
| GPT-5.4 nano (`gpt-5.4-nano-2026-03-17`) | Same reconstructed SBS protocol | Completed timestamped API run |
| GPT-5.6 Luna (`gpt-5.6-luna`) | Same reconstructed SBS protocol | Completed timestamped API run |

: Principal SBS model-generation comparison {#tbl-principal-model-comparison}

Plain-zero-shot and `improved-v1` artifacts remain preserved but are not part of the principal design. The archival `improved-v1` runner is dry-run by default, runs the current-model arm independently unless another arm is explicitly selected, and requires a separate `ARCHIVAL-GPT4` confirmation token before any GPT-4 Turbo API call. Such archival reruns write only to new run-specific directories and never replace frozen 2024 outputs.

## Cases and source documents

```{r}
#| label: tbl-cases
#| tbl-cap: "PONs in the strict SBS comparison"

tibble::tribble(
  ~PON, ~File,
  "IRG", "BO2A.txt",
  "EUHC", "BO3A.txt",
  "COAS", "BO4A.txt",
  "CAN", "BO5A.txt",
  "CEER", "BO6A.txt"
) |>
  knitr::kable()
```

BEREC is not pooled into this comparison. Dedicated BEREC replication uses `BO1B.txt` and separate guarded scripts.

## Reconstructed 2024 SBS protocol

The final Dec. 24 implementation used Chat Completions with:

1. a **system message** containing the policy-analyst instruction, the SBS suffix, and the ordered questions; and
2. a **user message** containing the complete boundary-object text.

The request supplied `model` and `messages` explicitly. No temperature, top-p, token limit, seed, response format, or tools were explicitly supplied in the historical code.

The recovered base instruction is:

> You are a policy analyst focused on extracting organizational features from documents that regulate the organization and functioning of specific organizations. I will provide the documents and a list of yes-no questions. You will only return TRUE or FALSE responses according to the list of yes-no questions, including a row with the question itself. The output must be in CSV format.

The manuscript-facing SBS condition appends:

> Let's think step by step

The project preserves the recovered string-concatenation behavior, including its whitespace quirks. `scripts/00_check_protocol.py` prints the resulting prompt hash without making an API call.

## Fifteen computational questions and fourteen manuscript prompts

```{r}
#| label: tbl-questions
#| tbl-cap: "Binary governance questions in the preserved computational instrument"

questions |>
  knitr::kable(col.names = c("Computational ID", "Question"))
```

The computational instrument contains 15 questions. Manuscript Table A5 contains 14 because the working-groups item—computational question 10—is omitted. The project does not delete or renumber the historical computational instrument. Instead it defines a separate 14-question manuscript subset for Table A5 replication.

## Human coders and majority vote

The authoritative human file is `input/reference/legacy_human_responses_confirmed.csv`. The individual coders are displayed as **HBM**, **ASC**, and **JNG**.

For each question/PON combination, the pipeline records:

- each individual human response;
- the three-human majority vote, used as the confirmed human reference;
- whether the human coders are unanimous; and
- each model classification and whether it agrees with that human majority.

Historical four-coder agreement is expressed with explicit counts such as `3 TRUE / 1 FALSE` rather than the manuscript's shorthand. The original `1`, `0`, `3-1`, `2-2`, and `3-1*` values are retained only as manuscript-reference fields.

## Frozen historical benchmark

The five `*_sbs.txt` files from Dec. 24, 2024 are treated as immutable research artifacts. The principal runner refuses `gpt-4-turbo` and `gpt-4-turbo-2024-04-09`, preventing accidental regeneration through the contemporary-model workflow.

## Contemporary-model runs

The principal comparison registers two contemporary arms: the pinned GPT-5.4 nano snapshot `gpt-5.4-nano-2026-03-17` and GPT-5.6 Luna using the currently exposed API model ID `gpt-5.6-luna`. Each arm was executed independently with `scripts/03_run_current_model.py`, using five requests—one per strict-comparison PON—and writing only to its own timestamped directory under `output/current/`. Completed runs are discovered and reused, so the current comparison does **not** rerun GPT-5.4 nano, GPT-5.6 Luna, or any historical output.

Each response receives a JSON metadata sidecar. A fully successful run receives `run_manifest.json` with `status: complete`; if execution fails after the run directory is created, partial outputs are retained and `run_failed.json` records the failure. Downstream comparison scripts accept only runs whose completion manifest explicitly reports `status == "complete"` and require one completed run for every model registered in `comparison.comparison_models`. OpenAI's GPT-5.6 Luna model page was checked on **2026-08-31**: its **Snapshots** section listed `gpt-5.6-luna` but no dated snapshot identifier. The project therefore treats `gpt-5.6-luna` as an **undated model ID**, records the documentation URL and check date in configuration, and preserves the UTC execution time plus requested and API-returned model IDs in response metadata and the run manifest. It is not described as a dated or immutable snapshot.

## BEREC replication boundary

The original BEREC Dec. 23 request state is not recoverable with enough certainty to support an exact-replication claim. The project therefore distinguishes:

- exact replication: not defensible from surviving request provenance;
- model-family replication: still limited by protocol uncertainty; and
- protocol replication: feasible as a retrospective application of the recovered Dec. 24 SBS protocol to BEREC with GPT-5.4 nano.

The best-supported dated GPT-4 Turbo snapshot is `gpt-4-turbo-2024-04-09`. Dedicated BEREC scripts use `BO1B.txt`, default to dry-run, require both `--execute` and `--confirm-api-call BEREC`, plan one API request, and write only under `output/berec_replication/`.
