---
title: "2024 Zero-Shot Results"
subtitle: "Evaluation of the preserved December 2024 outputs"
editor: visual
---

```{r}
#| label: setup
#| include: false
library(dplyr)
library(ggplot2)
library(knitr)
library(purrr)
library(readr)
library(tibble)

source("R/analysis.R")
questions <- readr::read_csv(
  "input/reference/questions.csv",
  show_col_types = FALSE
)
human <- readr::read_csv(
  "input/reference/legacy_human_responses.csv",
  show_col_types = FALSE
)

files <- tibble::tribble(
  ~entity_slug, ~path,
  "irg", "output/BO2A-2024-12-24-2-52-18_irg_z.txt",
  "euhc", "output/BO3A-2024-12-24-2-53-29_euhc_z.txt",
  "coas", "output/BO4A-2024-12-24-2-53-56_coas_z.txt",
  "can", "output/BO5A-2024-12-24-2-54-17_can_z.txt",
  "ceer", "output/BO6A-2024-12-24-2-54-38_ceer_z.txt"
)

evaluation <- purrr::pmap_dfr(
  files,
  function(entity_slug, path) {
    evaluate_output(
      path = path,
      entity_slug = entity_slug,
      technique = "Zero-shot",
      questions = questions,
      human = human
    ) |>
      dplyr::mutate(pon = pon_label(entity_slug))
  }
)

metrics <- evaluation |>
  dplyr::group_by(.data$entity_slug, .data$pon) |>
  dplyr::group_modify(~ compute_metrics(.x)) |>
  dplyr::ungroup()
```

## Scope

This archival report retains only the five **zero-shot** response files from the final Dec. 24, 2024 pipeline for which a recovered human-reference vector is available. Historical step-by-step files remain preserved on disk but are no longer part of the analytical design.

BEREC is retained only as archival provenance and is excluded from the comparative experiment. All results on this page therefore use IRG, EUHC, COAS, CAN, and CEER.

## Performance by PON

```{r}
#| label: tbl-legacy-metrics
#| tbl-cap: "Performance of the preserved 2024 zero-shot outputs against the recovered human reference"

metrics |>
  dplyr::mutate(
    accuracy = scales::percent(.data$accuracy, accuracy = 0.1),
    precision = scales::percent(.data$precision, accuracy = 0.1),
    recall = scales::percent(.data$recall, accuracy = 0.1),
    specificity = scales::percent(.data$specificity, accuracy = 0.1),
    f1 = scales::percent(.data$f1, accuracy = 0.1)
  ) |>
  dplyr::select(
    .data$pon,
    .data$accuracy,
    .data$precision,
    .data$recall,
    .data$specificity,
    .data$f1,
    .data$tp,
    .data$fp,
    .data$fn,
    .data$tn
  ) |>
  knitr::kable(
    col.names = c(
      "PON", "Accuracy", "Precision", "Recall", "Specificity",
      "F1", "TP", "FP", "FN", "TN"
    )
  )
```

::: {.callout-tip appearance="simple"}
**Question:** What share of each PON's 15 classifications disagreed with the recovered human reference?
:::

```{r}
#| label: fig-legacy-disagreement
#| fig-cap: "Relative weight of disagreements with the recovered human reference in the preserved 2024 zero-shot outputs. Each bar represents the share of 15 decisions that disagreed."
#| fig-height: 4.5

plot_data <- metrics |>
  dplyr::arrange(.data$disagreement_rate)

ggplot2::ggplot(
  plot_data,
  ggplot2::aes(
    x = stats::reorder(.data$pon, .data$disagreement_rate),
    y = .data$disagreement_rate
  )
) +
  ggplot2::geom_col(
    width = 0.65,
    fill = "#0072B2"
  ) +
  ggplot2::geom_text(
    ggplot2::aes(
      label = scales::percent(.data$disagreement_rate, accuracy = 0.1)
    ),
    hjust = -0.15,
    size = 3.5
  ) +
  ggplot2::coord_flip() +
  ggplot2::scale_y_continuous(
    limits = c(0, 1),
    breaks = seq(0, 1, by = 0.2),
    labels = scales::label_percent(accuracy = 1),
    expand = ggplot2::expansion(mult = c(0, 0.08))
  ) +
  ggplot2::labs(
    x = NULL,
    y = "Share disagreeing with human reference"
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(panel.grid.minor = ggplot2::element_blank())
```

## Question-level discrepancies

```{r}
#| label: tbl-legacy-errors
#| tbl-cap: "Question-level disagreements in the preserved 2024 zero-shot outputs"

evaluation |>
  dplyr::filter(!.data$agreement) |>
  dplyr::select(
    .data$question_id,
    .data$pon,
    .data$llm_response,
    .data$human_response,
    .data$eval
  ) |>
  dplyr::arrange(.data$question_id, .data$pon) |>
  knitr::kable(
    col.names = c("Question", "PON", "LLM", "Human", "Type")
  )
```

The purpose of this page is archival. The six-scenario report evaluates how classifications vary across model choice, protocol, and their combination.
