---
title: "2024 Replication"
subtitle: "Frozen SBS outputs, human coders, and manuscript Table A5"
editor: visual
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(dplyr)
library(knitr)
library(readr)

replication_path <- "derived/table_a5_replication.csv"
discrepancy_path <- "derived/table_a5_discrepancies.csv"
# metrics_path <- "derived/legacy_sbs_human_metrics.csv"

metrics_path <- "derived/sbs_model_metrics_by_pon.csv"

replication_ready <- file.exists(replication_path)
metrics_ready <- file.exists(metrics_path)

if (replication_ready) {
  a5 <- readr::read_csv(replication_path, show_col_types = FALSE)
  discrepancies <- readr::read_csv(discrepancy_path, show_col_types = FALSE)
}
if (metrics_ready) {
  legacy_metrics <- readr::read_csv(
    metrics_path,
    show_col_types = FALSE
  ) |>
    dplyr::filter(model_id == "gpt-4-turbo")
}
```

## Historical object being replicated

The manuscript-facing historical comparison contains four coders: **HBM, ASC, JNG, and the frozen GPT-4 Turbo SBS response**. The manuscript reports 14 prompts over five PONs, producing 70 agreement cells.

The reconstruction script uses the confirmed individual human responses and the preserved Dec. 24 `*_sbs.txt` files. It never calls OpenAI.

``` bash
Rscript scripts/02_reconstruct_table_a5.R
Rscript scripts/02_evaluate_preserved_outputs.R
```

## Agreement terminology

The reconstructed project uses explicit response counts rather than making the reader decode `1`, `0`, `3-1`, or `2-2`. For example, `3 TRUE / 1 FALSE` states both the degree and direction of disagreement. A separate field identifies a single dissenting coder when applicable.

The project also reports whether the legacy LLM agrees with the **three-human majority vote**.

## Table A5 replication result

```{r}
#| label: a5-status
#| results: asis
if (!replication_ready) {
  cat("::: {.callout-important}\nRun `Rscript scripts/02_reconstruct_table_a5.R` to build the 70-cell replication audit.\n:::\n")
} else {
  cat(sprintf(
    "The surviving data reproduce **%s of %s** Table A5 cells. **%s** cells differ from the manuscript table.\n",
    sum(a5$table_a5_match), nrow(a5), sum(!a5$table_a5_match)
  ))
}
```

```{r}
#| label: tbl-a5-discrepancies
#| tbl-cap: "Table A5 cells that do not match the surviving raw human and frozen SBS data"
if (replication_ready) {
  discrepancies |>
    dplyr::select(
      entity_slug,
      manuscript_question_id,
      HBM,
      ASC,
      JNG,
      legacy_llm_sbs,
      historical_four_coder_pattern,
      manuscript_agreement,
      notes
    ) |>
    knitr::kable(
      col.names = c(
        "PON", "Prompt", "HBM", "ASC", "JNG", "Legacy LLM SBS",
        "Surviving pattern", "Table A5", "Conclusion"
      )
    )
}
```

The three discrepancies are IRG prompt 3, cOAlition S prompt 8, and CEER prompt 6. They are recorded as Table A5 errors; the source responses are not edited to make the table reproduce.

## Frozen SBS performance against the confirmed human reference

```{r}
#| label: tbl-legacy-metrics
#| tbl-cap: "Frozen 2024 GPT-4 Turbo SBS performance by PON over the 15-question computational instrument"
if (metrics_ready) {
  legacy_metrics |>
    dplyr::select(
      pon,
      n,
      accuracy,
      disagreement_rate,
      precision,
      recall,
      specificity,
      f1
    ) |>
    dplyr::mutate(
      accuracy = scales::percent(accuracy, accuracy = 0.1),
      disagreement_rate = scales::percent(
        disagreement_rate,
        accuracy = 0.1
      ),
      precision = scales::percent(precision, accuracy = 0.1),
      recall = scales::percent(recall, accuracy = 0.1),
      specificity = scales::percent(specificity, accuracy = 0.1),
      f1 = scales::percent(f1, accuracy = 0.1)
    ) |>
    knitr::kable(
      col.names = c(
        "PON",
        "N",
        "Accuracy",
        "Disagreement",
        "Precision",
        "Recall",
        "Specificity",
        "F1"
      ),
      align = c("l", "r", "r", "r", "r", "r", "r", "r")
    )
}
```

This evaluation uses all 15 computational questions. The 14-question subset is used only where the project explicitly reconstructs manuscript Table A5.
