---
title: "2024 Replication"
subtitle: "Frozen SBS outputs, human coders, and manuscript Table A5"
editor: visual
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(dplyr)
library(knitr)
library(readr)

replication_path <- "derived/table_a5_replication.csv"
discrepancy_path <- "derived/table_a5_discrepancies.csv"
# metrics_path <- "derived/legacy_sbs_human_metrics.csv"

metrics_path <- "derived/sbs_model_metrics_by_pon.csv"

replication_ready <- file.exists(replication_path)
metrics_ready <- file.exists(metrics_path)

if (replication_ready) {
  a5 <- readr::read_csv(replication_path, show_col_types = FALSE)
  discrepancies <- readr::read_csv(discrepancy_path, show_col_types = FALSE)
}
if (metrics_ready) {
  legacy_metrics <- readr::read_csv(
    metrics_path,
    show_col_types = FALSE
  ) |>
    dplyr::filter(model_id == "gpt-4-turbo")
}
```

## Historical object being replicated

The manuscript-facing historical comparison contains four coders: **HBM, ASC, JNG, and the frozen GPT-4 Turbo SBS response**. The manuscript reports 14 prompts over five PONs, producing 70 agreement cells.

The reconstruction script uses the confirmed individual human responses and the preserved Dec. 24 `*_sbs.txt` files. It never calls OpenAI.

``` bash
Rscript scripts/02_reconstruct_table_a5.R
Rscript scripts/02_evaluate_preserved_outputs.R
```

## Agreement terminology

The reconstructed project uses explicit response counts rather than making the reader decode `1`, `0`, `3-1`, or `2-2`. For example, `3 TRUE / 1 FALSE` states both the degree and direction of disagreement. A separate field identifies a single dissenting coder when applicable.

The project also reports whether the legacy LLM agrees with the **three-human majority vote**.

## Table A5 replication result

```{r}
#| label: a5-status
#| results: asis
if (!replication_ready) {
  cat("::: {.callout-important}\nRun `Rscript scripts/02_reconstruct_table_a5.R` to build the 70-cell replication audit.\n:::\n")
} else {
  cat(sprintf(
    "The surviving data reproduce **%s of %s** Table A5 cells. **%s** cells differ from the manuscript table.\n",
    sum(a5$table_a5_match), nrow(a5), sum(!a5$table_a5_match)
  ))
}
```

```{r}
#| label: tbl-a5-discrepancies
#| tbl-cap: "Table A5 cells that do not match the surviving raw human and frozen SBS data"
if (replication_ready) {
  discrepancies |>
    dplyr::select(
      entity_slug,
      manuscript_question_id,
      HBM,
      ASC,
      JNG,
      legacy_llm_sbs,
      historical_four_coder_pattern,
      manuscript_agreement,
      notes
    ) |>
    knitr::kable(
      col.names = c(
        "PON", "Prompt", "HBM", "ASC", "JNG", "Legacy LLM SBS",
        "Surviving pattern", "Table A5", "Conclusion"
      )
    )
}
```

The three discrepancies are IRG prompt 3, cOAlition S prompt 8, and CEER prompt 6. They are retained as Table A5 discrepancies relative to the surviving coder-level data; the source responses are not edited to force agreement with the manuscript table.

### Manuscript agreement findings supported by the surviving data

The manuscript highlights several agreement patterns that can be checked directly from the surviving five-PON coder-level data. The summaries below are computed from the same reconstructed 70-cell object rather than transcribed from the manuscript.

```{r}
#| label: manuscript-agreement-findings
#| results: asis
if (replication_ready) {
  agreement_findings <- a5 |>
    dplyr::mutate(
      full_agreement = rowSums(
        dplyr::across(
          c(HBM, ASC, JNG, legacy_llm_sbs),
          ~ .x == TRUE
        )
      ) %in% c(0L, 4L),
      PON = dplyr::recode(
        .data$entity_slug,
        irg = "IRG",
        euhc = "EUHC",
        coas = "cOAlition S",
        can = "CAN",
        ceer = "CEER"
      )
    )

  prompt_agreement <- agreement_findings |>
    dplyr::group_by(.data$manuscript_question_id, .data$question) |>
    dplyr::summarise(
      full_agreement_pons = sum(.data$full_agreement),
      total_pons = dplyr::n(),
      .groups = "drop"
    ) |>
    dplyr::arrange(.data$manuscript_question_id)

  pon_agreement <- agreement_findings |>
    dplyr::group_by(.data$PON) |>
    dplyr::summarise(
      full_agreement_prompts = sum(.data$full_agreement),
      total_prompts = dplyr::n(),
      .groups = "drop"
    ) |>
    dplyr::mutate(
      proportion = sprintf(
        "%s/%s (%.1f%%)",
        .data$full_agreement_prompts,
        .data$total_prompts,
        100 * .data$full_agreement_prompts / .data$total_prompts
      ),
      PON = factor(
        .data$PON,
        levels = c("IRG", "cOAlition S", "EUHC", "CEER", "CAN")
      )
    ) |>
    dplyr::arrange(.data$PON)

  unanimous_prompts <- prompt_agreement |>
    dplyr::filter(.data$full_agreement_pons == .data$total_pons) |>
    dplyr::pull(.data$manuscript_question_id)

  if (!identical(as.integer(unanimous_prompts), c(1L, 7L, 9L, 11L, 12L))) {
    stop("Unexpected set of fully agreeing manuscript prompts")
  }

  q5_agreement <- prompt_agreement |>
    dplyr::filter(.data$manuscript_question_id == 5L) |>
    dplyr::pull(.data$full_agreement_pons)

  if (!identical(as.integer(q5_agreement), 0L)) {
    stop("Unexpected prompt 5 full-agreement count")
  }

  cat(
    "Across the **five PONs represented in the surviving Table A5 data**, prompts **1, 7, 9, 11, and 12** show full four-coder agreement for all **5/5 PONs**. Prompt **5** is the opposite extreme: **0/5 PONs** show full four-coder agreement.\n\n"
  )
}
```

```{r}
#| label: tbl-prompt-agreement-counts
#| tbl-cap: "Number of PONs with full four-coder agreement for each manuscript prompt"
if (replication_ready) {
  prompt_agreement |>
    dplyr::transmute(
      Prompt = .data$manuscript_question_id,
      Question = .data$question,
      `Full agreement` = sprintf(
        "%s/%s",
        .data$full_agreement_pons,
        .data$total_pons
      )
    ) |>
    knitr::kable(align = c("r", "l", "r"))
}
```

```{r}
#| label: tbl-pon-agreement-counts
#| tbl-cap: "Number of manuscript prompts with full four-coder agreement by PON"
if (replication_ready) {
  pon_agreement |>
    dplyr::transmute(
      PON = as.character(.data$PON),
      `Full-agreement prompts` = .data$proportion
    ) |>
    knitr::kable(align = c("l", "r"))
}
```

EUHC and CEER have the highest full-agreement counts at **11/14 prompts (78.6%)** each, followed by IRG at **10/14 (71.4%)**, cOAlition S at **8/14 (57.1%)**, and CAN at **7/14 (50.0%)**. These are descriptive agreement summaries of the surviving five-PON historical comparison; they are not model-accuracy estimates.

### Full 70-cell reconstruction

The table below reports all 70 manuscript-facing cells reconstructed from the surviving individual human codings and frozen SBS responses. The **Table A5** column preserves the manuscript transcription; **Match** indicates whether that cell is reproduced by the surviving source data.

```{r}
#| label: tbl-a5-full-reconstruction
#| tbl-cap: "Full reconstruction of the 70 manuscript Table A5 cells"
if (replication_ready) {
  a5 |>
    dplyr::mutate(
      PON = dplyr::recode(
        .data$entity_slug,
        irg = "IRG",
        euhc = "EUHC",
        coas = "cOAlition S",
        can = "CAN",
        ceer = "CEER"
      ),
      Match = dplyr::if_else(.data$table_a5_match, "Match", "Mismatch")
    ) |>
    dplyr::arrange(
      factor(.data$PON, levels = c("IRG", "EUHC", "cOAlition S", "CAN", "CEER")),
      .data$manuscript_question_id
    ) |>
    dplyr::select(
      .data$PON,
      Prompt = .data$manuscript_question_id,
      .data$HBM,
      .data$ASC,
      .data$JNG,
      `Legacy LLM SBS` = .data$legacy_llm_sbs,
      `Surviving pattern` = .data$historical_four_coder_pattern,
      `Table A5` = .data$manuscript_agreement,
      .data$Match
    ) |>
    knitr::kable(
      align = c("l", "r", "l", "l", "l", "l", "l", "l", "l")
    )
}
```

## Frozen SBS agreement with the confirmed human reference

```{r}
#| label: tbl-legacy-metrics
#| tbl-cap: "Frozen 2024 GPT-4 Turbo SBS agreement with the confirmed human reference by PON"
if (metrics_ready) {
  legacy_metrics |>
    dplyr::select(
      pon,
      n,
      accuracy,
      disagreement_rate,
      precision,
      recall,
      specificity,
      f1
    ) |>
    dplyr::mutate(
      accuracy = scales::percent(accuracy, accuracy = 0.1),
      disagreement_rate = scales::percent(
        disagreement_rate,
        accuracy = 0.1
      ),
      precision = scales::percent(precision, accuracy = 0.1),
      recall = scales::percent(recall, accuracy = 0.1),
      specificity = scales::percent(specificity, accuracy = 0.1),
      f1 = scales::percent(f1, accuracy = 0.1)
    ) |>
    knitr::kable(
      col.names = c(
        "PON",
        "N",
        "Agreement",
        "Disagreement",
        "Precision",
        "Recall",
        "Specificity",
        "F1"
      ),
      align = c("l", "r", "r", "r", "r", "r", "r", "r")
    )
}
```

This evaluation uses all 15 computational questions. The 14-question subset is used only where the project explicitly reconstructs manuscript Table A5. Precision, recall, specificity, and F1 treat the three-human majority classification as the reference label; they are descriptive reference-based diagnostics rather than claims of external ground-truth accuracy.
